\coloc: Optimize Spatial Visualization Following Gestalt Principles

### 2023-03-17 Manuscript/manuscript.tex
add references.

%% INTRODUCTION ============================================================================================
\section*{Introduction}
Visualization is an indispensable component of data analysis, providing clarity that connects quantitative evidence to key conclusions. \cite{dagostinomcgowan_2022} In bioinformatics research, visualization receives growing recognition as essential: many scientists rely on visualization to complete their cognitive process from analysis to insight, including analytic validation of automated pipelines and scientific communication. \cite{odonoghue_2021} Nevertheless, current methods and tools are inadequate to visualize the growingly complex biological data and computational models, limiting the ability to derive new insights and discoveries. \cite{odonoghue_2010}

The recent technology breakthrough in spatial profiling enables characterizing the diversity of molecules, e.g. DNA, RNA, and proteins, in the spatial context \cite{moffitt_2022}, generating novel spatially-resolved data that requires highly tailored \textit{in situ} visualizations \cite{dries_2021, Lewis_2021, odonoghue_2021}. Current \textit{in situ} visualizations, as opposed to embedding visualizations (e.g. t-SNE and UMAP) that project information to some mathematical space, visualize molecular information in the original cellular location, creating spatial maps that color-code a single variable (e.g., molecule expression, cell type, or spatial domain). This approach lacks the ability to simultaneously exhibit multiple variables, possibly from disparate data domains (such as expression domain and spatial domain) or modalities (such as transcriptomics and proteomics), creating cognitive gaps to interpret findings regarding their (micro-) environment. For example, when studying differential expression in different spatial domains (commonly defined by tissue architecture, e.g. cancerous regions, cortical layers), two spatial maps are displayed side-by-side, one for gene expression and one for spatial domains, creating significant challenges to associate these two information together. While software-based interactive visualizations or 3-D visualizations have the potential to mitigate this challenge, they are infeasible for scientific communications in static media, such as prints. Developing a static spatial visualization that enables the simultaneous display of multiple dimensions of information is crucial for spatially-resolved data research.

@@

 To optimize \textit{in situ} visualization, we introduce gestalt principles to the design of spatial maps. Gestalt principles \cite{todorovic_2008, palmer_1999} refer to a set of rules describing how humans perceive and interpret visual information and are commonly applied in art and designs. Following gestalt principles, we add additional visual dimensions to spatial maps, leveraging different aesthetics, such as color, fill, and symbols, to simultaneously display disparate variables. Specifically, we apply figure-ground articulation to display two variables: one variable (e.g. expression) can be plotted as color-filled circles, serving as the \textit{figure}; one variable (e.g. spatial domains) can be plotted as the backgrounds of the circles, creating a \textit{ground} for the figure. For adjacent circles with limited space between them to display the background color, we use an economic implementation, colored outlines for these circles, inspired by watercolor effect \cite{pinna_1987, pinna_2001}. Watercolor effect describes the phenomenon in visual perception that surface color arises from thin boundaries and hence is applied here to perceive the background color in tight space. Overall, the figure-ground articulation creates two isolated layers in visual perception to display the two variables while maintaining the relative spatial relationship serving as a reference between the two. In addition, other fundamental principles \cite{todorovic_2008}, such as proximity, similarity, continuity, and closure, incentivize the brain to group elements and dimensions in the visualization, guaranteeing an integrative perception of the complex multi-dimensional spatial map.

To implement the proposed visualization, we leverage the state-of-art data visualization framework \texttt{ggplot2} \cite{ggplot2} following the grammar of graphics \cite{wilkinson_2012}. Specifically, we use an interactive process to layer each variable in spatial maps using different aesthetics. For example, we use the combination of \texttt{color} and \texttt{fill="transparent"} to create the background layer and \texttt{fill} to create the figure layer. When necessary to display an additional layer for a third variable, \texttt{shape} can be used to add symbols such as cross (+) and asterisk (*) to highlight in the spatial map. We provide an open-source package called \coloc (named after the graphic artist M.C. Escher) in the R programming environment \cite{R}, providing a simplified interface to navigate the implementation of the multi-dimensional \textit{in situ} visualization. By adapting \texttt{ggplot2} standard, \coloc can be seamlessly integrated into many popular spatial resolved toolboxes, such as \texttt{SpatialLIBD}\cite{pardo_2022}, \texttt{Seurat}\cite{hao_2021}, \texttt{Giotto} \cite{dries_2021} to name a few, and allow further theme customization with ease. We provide two use cases to examplify some utility of the proposed spatial visualization: 1) the spatially differential gene colocalization in the human dorsolateral prefrontal cortex (DLPFC) using spatial transcriptomics data; 2) multi-dimensional UMAP highlighting differential gene expression in data-driven cell clusters.

>>>>

%% INTRODUCTION ============================================================================================
\section*{Introduction}
Visualization is an indispensable component of data analysis, providing clarity that connects quantitative evidence to key conclusions.\cite{dagostinomcgowan_2022} In bioinformatics research, visualization receives growing recognition as essential: many scientists rely on visualization to complete their cognitive process from analysis to insight, including analytic validation of automated pipelines and scientific communication.\cite{odonoghue_2021} Nevertheless, current methods and tools are inadequate to visualize the growingly complex biological data and computational models, limiting the ability to derive new insights and discoveries.\cite{odonoghue_2010}

The recent technology breakthrough in spatial profiling enables characterizing the diversity of molecules, e.g. DNA, RNA, and proteins, in the spatial context\cite{moffitt_2022}, generating novel spatially-resolved data that requires highly tailored \textit{in situ} visualizations\cite{dries_2021, Lewis_2021, odonoghue_2021}. Current \textit{in situ} visualizations, as opposed to embedding visualizations (e.g. t-SNE and UMAP) that project information to some mathematical space, visualize molecular information in the original cellular location, creating spatial maps that color-code a single variable (e.g., molecule expression, cell type, or spatial domain). This approach lacks the ability to simultaneously exhibit multiple variables, possibly from disparate data domains (such as expression domain and spatial domain) or modalities (such as transcriptomics and proteomics), creating cognitive gaps to interpret findings regarding their (micro-) environment. For example, when studying differential expression in different spatial domains (commonly defined by tissue architecture, e.g. cancerous regions, cortical layers), two spatial maps are displayed side-by-side, one for gene expression and one for spatial domains, creating significant challenges to associate these two information together. While software-based interactive visualizations\cite{sriworarat_2023} or 3-D visualizations have the potential to mitigate this challenge, they are infeasible for scientific communications in static media, such as prints. Developing a static spatial visualization that enables the simultaneous display of multiple dimensions of information is crucial for spatially-resolved data research.

@@

 To optimize \textit{in situ} visualization, we introduce gestalt principles to the design of spatial maps. Gestalt principles\cite{todorovic_2008, palmer_1999} refer to a set of rules describing how humans perceive and interpret visual information and are commonly applied in art and designs. Following gestalt principles, we add additional visual dimensions to spatial maps, leveraging different aesthetics, such as color, fill, and symbols, to simultaneously display disparate variables. Specifically, we apply figure-ground articulation to display two variables: one variable (e.g. expression) can be plotted as color-filled circles, serving as the \textit{figure}; one variable (e.g. spatial domains) can be plotted as the backgrounds of the circles, creating a \textit{ground} for the figure. For adjacent circles with limited space between them to display the background color, we use an economic implementation, colored outlines for these circles, inspired by watercolor effect\cite{pinna_1987, pinna_2001}. Watercolor effect describes the phenomenon in visual perception that surface color arises from thin boundaries and hence is applied here to perceive the background color in tight space. Overall, the figure-ground articulation creates two isolated layers in visual perception to display the two variables while maintaining the relative spatial relationship serving as a reference between the two. In addition, other fundamental principles\cite{todorovic_2008}, such as proximity, similarity, continuity, and closure, incentivize the brain to group elements and dimensions in the visualization, guaranteeing an integrative perception of the complex multi-dimensional spatial map.

To implement the proposed visualization, we leverage the state-of-art data visualization framework \texttt{ggplot2}\cite{ggplot2} following the grammar of graphics\cite{wilkinson_2012}. Specifically, we use an interactive process to layer each variable in spatial maps using different aesthetics. For example, we use the combination of \texttt{color} and \texttt{fill="transparent"} to create the background layer and \texttt{fill} to create the figure layer. When necessary to display an additional layer for a third variable, \texttt{shape} can be used to add symbols such as cross (+) and asterisk (*) to highlight in the spatial map. We provide an open-source package called \coloc (named after the graphic artist M.C. Escher) in the R programming environment\cite{R}, providing a simplified interface to navigate the implementation of the multi-dimensional \textit{in situ} visualization. By adapting \texttt{ggplot2} standard, \coloc can be seamlessly integrated into many popular spatial resolved toolboxes, such as \texttt{SpatialLIBD}\cite{pardo_2022}, \texttt{Seurat}\cite{hao_2021}, \texttt{Giotto}\cite{dries_2021} to name a few, and allow further theme customization with ease. We provide two use cases to exemplify some utility of the proposed spatial visualization: 1) the spatially differential gene colocalization in the human dorsolateral prefrontal cortex (DLPFC) using spatial transcriptomics data\cite{huukimyers_2023}; 2) multi-dimensional UMAP highlighting differential gene expression in data-driven cell clusters \cite{freytag_2020}.
